---
title       : Brief Introduction to R
subtitle    : Why you should learn this "language"?
author      : David Yang
job         : R Enthusiast
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
github      :
  user      : saluteyang
  repo      : slidifyDemo
---

## Agenda

1. Introduce some basic concepts of the statistical computing language
2. Introduce two packages that are powerful and popular
3. Point to further uses of R with a fun example

---

## This is easier than you think:

> 1. R is a high level language that skips over most programming niceties
> 2. R has a big support community and more packages than you have time to explore
> 3. Anything you want to achieve, someone probably has done it in R, and done it beautifully

---

## Let's start with `ggplot2`

> * Contributed by Hadley Wickham of Rice University
> * Some data to start us off
> * Historical price-load relationship during August based on hourly PJM data
```{r echo=F, fig.width=5, fig.height=5,results='asis',message=F,cache=T}
require(xlsx)
require(ggplot2)
priceload<-read.csv("C:\\Users\\Yangster\\Dropbox\\Private\\3AugHrlyLdPrice_PJM.csv")
p<-ggplot(priceload,aes(x=E_W_LD,y=Price))+ geom_point()+xlab("Load")+
  ylab("Price")
p+facet_grid(Date~.)
p+facet_grid(Date~.)+geom_smooth()
```
---

## What happened

```{r echo=T, eval=F}
require(xlsx)
require(ggplot2)
priceload<-read.csv("C:\\Users\\Yangster\\Dropbox\\Private\\3AugHrlyLdPrice_PJM.csv")
p<-ggplot(priceload,aes(x=E_W_LD,y=Price))+ geom_point()+xlab("Load")+
  ylab("Price")
p+facet_grid(Date~.)
p+facet_grid(Date~.)+geom_smooth()
```

With three simple lines of code, we
* Imported data from an Excel file
* Created elegant charts using `ggplot2`'s graphics language
  * using aesthetics (aes) to create mapping of data to graphic elements
  * using geometry (geom) to specify the way we want to present the data including sophisticated "faceting" and non-parametric regression

---

## Data visualization done right

Let's look at another example using historical forward data for Henry Hub natural gas for August 2013 delivery. Trade dates of each month are colored differently, a local regression line is produced as well as it's error bands.

```{r echo=F,fig.align='center',fig.width=7,fig.height=5,results='asis',message=F,cache=TRUE}
require(xlsx)
require(lubridate)
require(ggplot2)
futNGDel= read.xlsx("C:\\Users\\Yangster\\Dropbox\\Private\\FUTURE_NG_Del.xlsx",2,header=TRUE)
s2<-ggplot(data=subset(futNGDel,subset=year(DelMo)==2013 & 
                        month(DelMo)==8 & TradeDate>as.Date('4/1/2013','%m/%d/%Y')),
          aes(x=TradeDate,y=Price,colour=as.factor(month(TradeDate))))
s2+geom_point()+geom_smooth()+theme(legend.position="none")
```

---

## Some programming asides

```{r echo=T, eval=F}
require(xlsx)
require(lubridate)
require(ggplot2)
futNGDel= read.xlsx("C:\\Users\\Yangster\\Dropbox\\Private\\FUTURE_NG_Del.xlsx",2,header=TRUE)
s2<-ggplot(data=subset(futNGDel,subset=year(DelMo)==2013 & 
                        month(DelMo)==8 & TradeDate>as.Date('4/1/2013','%m/%d/%Y')),
          aes(x=TradeDate,y=Price,colour=as.factor(month(TradeDate))))
s2+geom_point()+geom_smooth()+theme(legend.position="none")
```

> * We do need to mention some R-related programming knowledge such as the following
    * How to subset data
    * How to use `month()` and `year()` functions in the lubridate package to extract information from dates
    * How to convert dates stored as `factors` to dates stored as `dates` and back

---

## A lot more cool stuff awaits

Once we are more familiar with R, we can do a lot by typing relatively little. What's more important is probably the dynamic and reproducible nature of it.

Any guesses what the code below will do?

```{r echo=T, eval=F}
require(xlsx)
require(lubridate)
require(ggplot2)
# Importing and processing of data
price.Seg= read.csv("C:\\Users\\Yangster\\Dropbox\\Private\\3AugHrlyLdPrice_PJM_2.csv",header=TRUE)
price.Seg$Date_Long<-as.Date(price.Seg$Date_Long,"%m/%d/%Y")
price.Seg<-price.Seg[,c('Date_Long','Date','Price','SegType')]
aggreg.Seg<-aggregate(price.Seg,list(price.Seg$SegType,price.Seg$Date_Long),mean)
aggreg.Seg<-aggreg.Seg[,c(1,2,4,5)]
names(aggreg.Seg)<-c('peaktype','date','year','price')
```

---

## (continued from above)

```{r echo=T, eval=F}
aggreg.Seg<-aggreg.Seg[order(aggreg.Seg$peaktype,aggreg.Seg$date),]
aggreg.Seg$peaktype<-factor(aggreg.Seg$peaktype,levels=c('OnPeak','OffPeak',"Wknd"))
# Setting up plotting
p_seg<-ggplot(aggreg.Seg,aes(x=date,y=price,colour=factor(year)))+
  geom_point()+xlab("Date")+ylab("Price")
p_seg+facet_grid(peaktype~year,scale='free_x')+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5))
p_seg_hist<-ggplot(aggreg.Seg,aes(x=price,fill=factor(year),colour=factor(year),..density..))+
  geom_density(alpha=0.5,position="identity")
p_seg_hist+facet_grid(peaktype~.,scale='free_x')
```

---

## Side-by-side comparison of August power prices by segment and year

```{r echo=F,fig.width=6.5,fig.height=6.5,results='asis',message=F,warning=FALSE,cache=TRUE}
require(xlsx)
require(lubridate)
require(ggplot2)
# Importing and processing of data
price.Seg= read.csv("C:\\Users\\Yangster\\Dropbox\\Private\\3AugHrlyLdPrice_PJM_2.csv",header=TRUE)
price.Seg$Date_Long<-as.Date(price.Seg$Date_Long,"%m/%d/%Y")
price.Seg<-price.Seg[,c('Date_Long','Date','Price','SegType')]
aggreg.Seg<-aggregate(price.Seg,list(price.Seg$SegType,price.Seg$Date_Long),mean)
aggreg.Seg<-aggreg.Seg[,c(1,2,4,5)]
names(aggreg.Seg)<-c('peaktype','date','year','price')
aggreg.Seg<-aggreg.Seg[order(aggreg.Seg$peaktype,aggreg.Seg$date),]
aggreg.Seg$peaktype<-factor(aggreg.Seg$peaktype,levels=c('OnPeak','OffPeak',"Wknd"))
# Setting up plotting
p_seg<-ggplot(aggreg.Seg,aes(x=date,y=price,colour=factor(year)))+
  geom_point()+xlab("Date")+ylab("Price")
p_seg+facet_grid(peaktype~year,scale='free_x')+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5))
p_seg_hist<-ggplot(aggreg.Seg,aes(x=price,fill=factor(year),colour=factor(year),..density..))+
  geom_density(alpha=0.5,position="identity")
p_seg_hist+facet_grid(peaktype~.,scale='free_x')
```

---

## Now let's look at `forecast`

> * Contributed by Rob Hyndman of Monash University, Australia
> * An excellent source of information on time series modeling using R is here: https://www.otexts.org/fpp/resources
> * The package `forecast` enables complex methods to be invoked with just the right function call
> * We'll see an example using a double-seasonality Holt-Winters procedure to forecast hourly power prices using the same hourly PJM data
    * The double seasonality refers to both the daily cycle as well as the weekly (weekday/weekend) cycle
> * The example uses a training set of 600 hourly data; the forecast (blue) is compared to the actual out-of-sample data of hour number 601 to 744 (plotted in black)

---

## Automated time-series forecast

```{r echo=F, results='asis', message=F,warning=FALSE,cache=TRUE}
require(forecast)
test<-price.Seg[price.Seg$Date==2012,]$Price
test.train<-test[1:600]
test.rest<-test[601:length(test)]
test.train<-ts(test.train,start=1,deltat=1)
fcast.test<-dshw(test.train, period1=24,period2=24*7)
require(rCharts)
a1<-test
a2=c(test.train,fcast.test$mean[1:144])
temp12<-seq(as.POSIXct("2012-08-01 01:00:00"), as.POSIXct("2012-09-01 00:00:00"), by="hour")
temp12.dat<-cbind.data.frame(temp12,a1,a2)
names(temp12.dat)<-c('Date','Actual','Forecast')
temp12.dat2<-transform(temp12.dat,Date=as.character(Date))
m2 <- mPlot(x = 'Date', y = c('Actual','Forecast'), type = 'Line',
            data = temp12.dat2, ymax=120, lineColors=c('#1E90FF','#000000'))
m2$set(pointSize = 0, lineWidth = 2)
m2
```

---

## Automated time-series forecast

```{r echo=F,results='asis',message=F,warning=FALSE,cache=TRUE}
test<-price.Seg[price.Seg$Date==2013,]$Price
test.train<-test[1:600]
test.rest<-test[601:length(test)]
test.train<-ts(test.train,start=1,deltat=1)
fcast.test<-dshw(test.train, period1=24,period2=24*7)
a1<-test
a2=c(test.train,fcast.test$mean[1:144])
temp13<-seq(as.POSIXct("2013-08-01 01:00:00"), as.POSIXct("2013-09-01 00:00:00"), by="hour")
temp13.dat<-cbind.data.frame(temp13,a1,a2)
names(temp13.dat)<-c('Date','Actual','Forecast')
temp13.dat2<-transform(temp13.dat,Date=as.character(Date))
m3 <- mPlot(x = 'Date', y = c('Actual','Forecast'), type = 'Line',
            data = temp13.dat2, ymax=80, lineColors=c('#1E90FF','#000000'))
m3$set(pointSize = 0, lineWidth = 2)
m3
```

---

## Let's have some fun, finally

Here is a neat way of visualizing 10 paths of random walk. This is realized in R using the googleVis package.
```{r, message=FALSE,echo=FALSE}
library(googleVis)
op <- options(gvis.plot.tag='chart')
require(reshape2)
require(googleVis)
s_0=10
mu=0.05/250
sigma=0.2/sqrt(250)
set.seed(123)
x=matrix(s_0,ncol=10,nrow=50) 
for (t in 2: 50)                 
  x[t,]=x[t-1,]+rnorm(10)*sigma+mu 
colnames(x)<-paste('Path',seq(1,10,by=1),sep="_")
new_x<-melt(x)
# Note that cbind alone will convert Date objects, use the following syntax instead
# Note also that cbind recycles the variable of shorter length
idx.Time<-seq.Date(as.Date('2011-01-01'),le=50,by='1 day')
x_chart<-cbind.data.frame(idx.Time,new_x)
colnames(x_chart)<-c("time","idx","path","price")
```

```{r, results='asis',echo=F}
BB.motion<-gvisMotionChart(x_chart,idvar="path",timevar="time",
                    xvar="time",yvar="price",options = list(width = 600, height = 450))
plot(BB.motion,"chart")
## Set options back to original options
options(op)
```
---